---
title: "DE with limma + KSEA"
output: html_notebook
---


```{r}
setwd("~/Documents/MSc Genomics Data Science/Research Project (ML)/Project Data")
```

```{r}
# Opening file
df.phosphoprot <- read.csv("DSS_imp_phospho_clusters.csv")

# Viewing the file
head(df.phosphoprot)
```

Data normalization


```{r}
# Log-transforming and normalizing the data

# Data distribution before normalization
boxplot(df.phosphoprot[,2:ncol(df.phosphoprot)],las=3, main="Data distribution before normalization")

# Log2 transform and normalize data by scaling
df.norm <- data.frame(site=df.phosphoprot$sites, 
                      scale((log2(df.phosphoprot[,2:ncol(df.phosphoprot)]))))

# Data distribution after normalization
#boxplot(df.norm[,2:ncol(df.norm)],las=3, main="Data distribution after Log transformation and normalization")

df.norm
```

Statistical Analysis - limma function
```{r}

install.packages("BiocManager")
library(BiocManager)
BiocManager::install("limma")
library(limma)
BiocManager::install("edgeR")
library(edgeR)
```



```{r}

compare.clusters.by.limma <- function(df.to.compare, log.data=TRUE){
  
  # Compare by limma
  #
  # first column is ppsite names
  # first set of samples are control
  # second set of samples are the test
  
  nc <- ncol(df.to.compare)
  df.s <- df.to.compare[,2:nc]
  
  if (log.data==TRUE | log.data==T){ 
    
    df.s <- scale(log2(df.s))
   
    }

  df.s1 <- data.frame(outcome=matrix(nrow=(36)))
  df.s2 <- data.frame(outcome=matrix(nrow=(25)))

  
  df.s1$outcome <- "cluster_1"
  df.s2$outcome <- "cluster_2"

  
  df.ss <- rbind(df.s1,df.s2)
   
  des <- factor(ifelse(df.ss$outcome=="cluster_1" ,"1",
                       "2"))


  facna <- addNA(des)
  design <- model.matrix(~ 0+factor(c(facna)))
  
  
  colnames(design) <- c("cluster_1","cluster_2")
  contrast.matrix <- makeContrasts(
                    cluster_1 - cluster_2,
                    levels=design) 
  

  fit <- lmFit(df.s,design)
  fit2 <- contrasts.fit(fit, contrast.matrix)
  fit2 <- eBayes(fit2)
  pvals <- data.frame(fit2$p.value)
  fvals <- data.frame(fit2$coefficients)


  df.xx <- data.frame(protein=df.to.compare[,1],
                      differnces=fvals,
                      pvalues=pvals)
  colnames(df.xx) <- c(colnames(df.to.compare)[1],"difference.cluster_1.vs.cluster_2","pvalues")
  
 
df.xx$qval <- p.adjust(df.xx$pvalues, method = "fdr")
  
  df.xx[,1] <-
    
  return(df.xx) 
}  
```

Running limma function  and sacing the results
```{r}

df.limma.results <- compare.clusters.by.limma(df.to.compare = df.norm,log.data = F)

write.csv(df.limma.results, "Results of limma analysis of phopho Cluster 1 vs Cluster 2.csv", row.names = F)
```



# Viewing the top most significant hits:
```{r}

head(df.limma.results[order(df.limma.results$difference.cluster_1.vs.cluster_2),])

# Select and view significant hits
df.significant <- subset(df.limma.results,df.limma.results$qval<0.11)
df.significant.decreased <- subset(df.significant,df.significant$difference.cluster_1.vs.cluster_2<0)
df.significant.increased <- subset(df.significant,df.significant$difference.cluster_1.vs.cluster_2>0)

nrow(df.significant.decreased)
nrow(df.significant.increased)

df.significant.decreased
df.significant.increased

df.significant.increased[order(df.significant.increased$pvalues), ]
df.significant.decreased[order(df.significant.decreased$pvalues), ]
```

Data visualization

```{r}
library(ggplot2)

pplot <- ggplot(df.limma.results,aes(x=difference.cluster_1.vs.cluster_2,y=-log10(qval)))+
  geom_point()+
  geom_point(data=df.significant.decreased,
             aes(x=difference.cluster_1.vs.cluster_2,y=-log10(qval)),color="blue")+
  geom_point(data=df.significant.increased,
             aes(x=difference.cluster_1.vs.cluster_2,y=-log10(qval)),color="red")+
  labs(x="Diff.Cluster 1 vs Cluster 2",
       title = "Phosphoproteomics of Group 1 vs Group 2")
pplot
```

Get significantly upregulated sites 
```{r}
df.significant.increased[order(df.significant.increased$pvalues),]  

```


Kinase substrate enrichment (KSEA) analysis to identify changes in kinase activities

```{r}
# Reading the limma results data
df <- read.csv("Results of limma analysis of phopho Cluster 1 vs Cluster 2.csv")

df
```

```{r}
# KSEA function

get.ks.set <- function(dataset){

  ## data sets

  pdts <- "https://www.dropbox.com/s/86jfnayv0qa1n2q/pdts.csv?dl=1"
  psite <- "https://www.dropbox.com/s/eb1qoofz793f4tq/psite.csv?dl=1"
  signor <- "https://www.dropbox.com/s/alpbq880emz1z2t/signor.csv?dl=1"
  df.datasets <- data.frame(dataset.names = c("signor","pdts","psite"),
                            datasets=c(signor,pdts,psite))
  myfile <- as.character(df.datasets[df.datasets$dataset.names==dataset,2])
  if (length(myfile)==0){
    print (paste(dataset, "not found. Check spelling"))
    print ("Select one of these:")
    print(df.datasets$dataset.names)
  }
  df.out <- read.csv(myfile)
  return(df.out)
}

ksear.s <- function(df.fold, ks_db){

  # df.fold == dataset of fold changes or
  # ks_db == database of kinase-substrate relationships
  #       possibilities are "signor" "pdts", "pSite"
  df.fold[,1] <- gsub("..",");",df.fold[,1],fixed = T)
  df.fold[,1] <- gsub(".","(",df.fold[,1],fixed = T)
  values.all <- na.omit(as.numeric(subset(df.fold[,2], df.fold[,2]!=0)))
  nc <- ncol(df.fold)
  df.ks <- get.ks.set(ks_db)
  nr <- nrow(df.ks)
  zscores <- numeric(nr)
  pvalues <- numeric(nr)
  msites <- numeric(nr)
  kinases <- character(nr)
  allsites <- character(nr)
  r=1
  for (r in 1:nr) {
    mym <- df.ks[r,2]
    kinase <- df.ks[r,1]
    if (is.na(mym) == F){
      if(mym>2){
        substrates <- as.character(df.ks[r,3])
        ss <- c(unlist(strsplit(substrates,";")))
        start.time <- Sys.time()
        df.xx <- subset(df.fold,df.fold[,1] %in% paste(ss,";",sep=""))
        sites <-paste(unlist(df.xx[,1]),collapse=";")
        if (nrow(df.xx)>2){
         # df.xx$prot.group <- kinase
          #sites.x <- data.frame(site=df.xx[,1])
          #sites.x$kinase <- kinase
          #all.sites <- rbind(all.sites,sites.x)

            myvalues <- na.omit(as.numeric(subset(df.xx[,2],df.xx[,2]!=0)))
            pval <- 1
            tryCatch({
              myks <- ks.test(values.all,myvalues,exact=F)
              pval <- myks$p.value
            }, error=function(e){}
            )
            mysd <- sd(values.all,na.rm = T)
            mymedian <- median(myvalues, na.rm = T)
            mymedian.all <- median(values.all,na.rm = T)
            sd.all <- sd(values.all)
            msites[r] <- length(myvalues)
            zscores[r] <- ((mymedian-mymedian.all)*((sqrt(msites[r])))/mysd)
            pvalues[r] <- pval
            kinases[r] <- kinase
            allsites[r] <- sites
          }
        }
      }
  }

  df.out <- data.frame(kinases,zscores,pvalues,m=msites,allsites)
  df.out <- subset(df.out,df.out$m>1)
  df.out$qval <- p.adjust(df.out$pvalues,method = "BH")
  return(df.out)
}
```


# Calling the ksear.s function to carry out KSEA
```{r}

df.k <- ksear.s(data.frame(sites=df$site, df$difference.cluster_1.vs.cluster_2),ks_db="pdts")


# View the top kinases $
df.k[order(df.k$pvalue),]

#df.k[order(df.k$pvalue),]
```


Enrichment in main kinases (based on positive z score)

Visualization of KSEA results

```{r}
# Selecting significant
k.selected <- subset(df.k,df.k$pvalues<.01)

# Drawing volcano plot and visualize it
plot.ksea.volcano <- ggplot(df.k,aes(x=zscores,y=-log10(qval)))+
  geom_point(aes(size=m))+
  geom_text(data = k.selected,aes(x=zscores,y=-log10(qval),label=kinases),hjust=-0.2)+
  labs(title = "KSEA of Cluster 1 vs Cluster 2")
plot.ksea.volcano

```

Getting significant kinases 

```{r}
significant_kinases<- df.k[df.k1$qval < 0.05, ]

significant_kinases <- significant_kinases[order(significant_kinases$qval), ]

significant_kinases <- head(significant_kinases, -1)
```



The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

