---
title: "R Notebook"
output: html_notebook
---

MissForest - missing data imputation using iterated random forests (Non-parametric)


Installing and loading the packages
```{r}
install.packages("missForest")
library(missForest)
install.packages("doParallel")
library(doParallel)
```

Setting a working directory 
```{r}
setwd("~/Documents/MSc Genomics Data Science/Research Project (ML)/Project Data")
```

Opening the file
```{r}
df.dss <- read.csv("DSS_Dataset.csv", row.names = 1)

df.dss
```

Removing random non-missing values and running MissForest to test the performance
```{r}
set.seed(123) 
```

Introducing random missing values
```{r}
df_copy <- df.dss.complete

missing_indices <- which(!is.na(df_copy), arr.ind = TRUE) # getting indices of non-missing values
num_missing <- floor(0.05 * length(missing_indices)) # choosing 5% of the data to be missing
```

Randomly selecting indices to be set as NA
```{r}
missing_sample <- missing_indices[sample(nrow(missing_indices), num_missing), ]
df_copy[missing_sample] <- NA
```

Perform missForest imputation
```{r}
num_cores <- detectCores() - 1  # Use one less than the total number of available cores
cl <- makeCluster(num_cores)
registerDoParallel(cl)

imputed_data <- missForest(
  xmis = df_copy,
  maxiter = 15,  # Maximum number of iterations
  ntree = 200,   # Number of trees in each forest
 # mtry = floor(sqrt(ncol(df_copy))),  # Adjust number of variables considered at each split
 # nodesize = c(5, 5),  # Set minimal size of terminal nodes for classification and regression
 # replace = FALSE,     # Experiment with sampling without replacement
 # parallelize = "forests",  # Enable parallel computation
  verbose = TRUE
)     
```

Getting the completed data 
```{r}
completed_data <- imputed_data$ximp
```

Viewing the OOB error
```{r}
imputed_data$OOBerror 
```

Comparing the imputed values with the original values
```{r}
original_values <- df.dss.complete[missing_sample]
imputed_values <- completed_data[missing_sample]

comparison <- data.frame(
  original = original_values,
  imputed = imputed_values
)

# printing the comparison
print(comparison)
```

Calculating RMSE and NRMSE
```{r}
rmse <- sqrt(mean((original_values - imputed_values)^2))

range_of_actuals <- max(original_values) - min(original_values)
nrmse <- rmse / range_of_actuals
```



Printing RMSE and NRMSE
```{r}
cat("Root Mean Squared Error (RMSE) of the imputation:", rmse, "\n")
cat("Normalized Root Mean Squared Error (NRMSE) of the imputation:", nrmse, "\n")
```

Calculating the R² value
```{r}
r_squared <- cor(comparison$original, comparison$imputed, use = "complete.obs")^2
cat("R² value:", r_squared, "\n")
```

Creating the scatter plot - model fit
```{r}
ggplot(comparison, aes(x = original, y = imputed)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Random Forest inputation: Actual vs Imputed",
       x = "Original Values",
       y = "Imputed Values") +
  theme_minimal() +
  annotate("text", x = Inf, y = -Inf, label = paste("R² = ", round(r_squared, 3)), 
           hjust = 1.1, vjust = -0.5, size = 5, color = "black")
```

Performing imputation
```{r}
imputed_data <- missForest(df.dss, ntree = 200, maxiter = 15, replace = TRUE)
completed_data <- imputed_data$ximp
```
Saving imputed data
```{r}
write.csv(completed_data, "imputed_DSS_data.csv")
```



